---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Rozo, Geoffrey"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

Once completed, you will "knit" and submit the resulting .html document and the .Rmd file. The .html will present the output of your R code and your written answers, but your R code will not appear.  Your R code will appear in the .Rmd file. The resulting .html document will be graded and a feedback report returned with comments.  Points assigned to each item appear in this template.

-----

The following code chunk will (a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine, (b) read-in the abalones dataset, defining a new data frame, "mydata," (c) return the structure of that data frame, and (d) calculate new variables, VOLUME and RATIO. If either package has not been installed, you must do so first via *install.packages()*; e.g. *install.packages("ggplot2")*. Do not include installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

# mydata <- read.csv(file.path("c:/Rabalone_new/", "abalones.csv"), sep = ",")
mydata <- read.csv("abalones.csv", sep = ",")

# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

(1)(a) (2 points) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a, eval = TRUE, echo = FALSE}
summary(mydata)

Class <- mydata$CLASS
Rings <- mydata$RINGS
Class_Rings <- table(Class, Rings)
Class_Rings

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: There are ten different variable types, with the Length, Diameter, and Height variables standing out to me with mild outliers on their lower ends, so I expect some left skewness. The variables Whole, Shuck, and Volume have mild outliers on their upper ends, and variables Rings and Ratio have extreme outliers, so I expect more right skewness for these variables. Also, variable Class A3 includes almost a third of the entire sample, so that may be important to keep in mind when comparing to small classes like A1 and A5.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
SEX <- factor(mydata$SEX, labels = c("Female", "Infant", "Male"))
CLASS <- mydata$CLASS
SEX_CLASS <- table(SEX, CLASS)
SEX_CLASS_SUM <- addmargins(SEX_CLASS)
SEX_CLASS_SUM

barplot(SEX_CLASS, beside=T, col=c("aquamarine3", "coral", "purple"), main = "CLASS membership, SEX-differentiated", xlab = "Class", ylab = "Frequency")

legend("topright", c("Female", "Infant", "Male"), pch = 15, col=c("aquamarine3", "coral", "purple"))

```

**Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: What stands out to me is that two-thirds of infant abalones are in the A1 and A2 classes alone, while over 80% of Female and Male abalones make up the A3, A4, and A5 classes. The A4 and A5 classes include only about 10% infant abalones, and the A1 class has barely any Female and Male abalones compared to infants.***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
work <- mydata[sample(1:nrow(mydata), 200, replace=FALSE),]
plot(work[, 2:6])

```

-----

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
ggplot(data = mydata, aes(x = mydata$VOLUME, y = mydata$WHOLE)) +
geom_point(aes(color = CLASS), size = 1.5) + ggtitle("Whole weight, in grams | Volume, cubic cm") +
labs(x = "Volume") + labs(y = "Whole weight")

```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
sw_ratio <- mydata$SHUCK/mydata$WHOLE    #first created a variable for Shuck/Whole from 'mydata' so I can use the 
#max as my slope in line 139.

ggplot(data = mydata, aes(x = mydata$WHOLE, y = mydata$SHUCK)) +
geom_point(aes(color = CLASS), size = 1.5) + ggtitle("Shuck weight, in grams | Whole weight, in grams") +
labs(x = "Whole weight") + labs(y = "Shuck weight") + geom_abline(lty = 2, slope = sw_ratio[which.max(sw_ratio)])
#largest sw_ratio is 0.5621 = slope

```

**Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: In this plot, classes seem to be less variable and grouped closer together than compared to the plot in (a) where classes A3, A4, and A5 are spread upwards more scattered and randomly. In this plot, when comparing to the largest Shuck/Whole ratio and looking at the three largest weighted classes, class A5 (purple) has the lowest Shuck/Whole ratio, class A4 (blue) is in the middle range ratio, and A3 (green) has the largest ratio. In both plots, the infant classes A1 and A2 do not show much variability staying near the low end of whole weight. ***

-----

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3, 3))

Female <- mydata[mydata$SEX %in% "F", ]    #searched through all entries of variable 'SEX' in 'mydata' to find the rows of all 'F' (Female) abalones

hist(Female$RATIO, col = rgb(.9, 0, 0), main = "Female RATIO", xlim = range(0, 0.30), xlab = " ")

Infant <- mydata[mydata$SEX %in% "I", ]    #searched through all entries of variable 'SEX' in 'mydata' to find the rows of all 'I' (Infant) abalones

hist(Infant$RATIO, col = rgb(0, .9, 0), main = "Infant RATIO", xlim = range(0, 0.30), xlab = " ")

Male <- mydata[mydata$SEX %in% "M", ]      #searched through all entries of variable 'SEX' in 'mydata' to find the rows of all 'M' (Male) abalones


hist(Male$RATIO, col = rgb(0, 0, .9), main = "Male RATIO", xlim = range(0, 0.30), xlab = " ")

boxplot(Female$RATIO, col = rgb(.9, 0, 0), main = "Female RATIO", ylim = range(0, 0.30), xlab = " ")

boxplot(Infant$RATIO, col = rgb(0, .9, 0), main = "Infant RATIO", ylim = range(0, 0.30), xlab = " ")

boxplot(Male$RATIO, col = rgb(0, 0, .9), main = "Male RATIO", ylim = range(0, 0.30), xlab = " ")

qqnorm(Female$RATIO, col = rgb(.9, 0, 0), main = "Female RATIO")
qqline(Female$RATIO)

qqnorm(Infant$RATIO, col = rgb(0, .9, 0), main = "Infant RATIO")
qqline(Infant$RATIO)

qqnorm(Male$RATIO, col = rgb(0, 0, .9), main = "Male RATIO")
qqline(Male$RATIO)

par(mfrow = c(1, 1))

```

**Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: All three histograms show right skewness and it can be seen from the boxplots that there are many outliers on the higher end of their ratios. The QQ plots also show a lot of deviation from the QQ lines, specifically on each plots upper end, again signifying right skewness. A normal distribution would have one large frequency bar in the histogram and equally smaller bars on each side, but the female and infant histograms have two equally large frequency bars, with smaller bars drifting to the right end with a very high outlier on the Female plot. The male histogram has two equally large bars on the right of the largest frequency bar and continues with more bars on the right side, so it also displays right skewness.***

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}
Female.Ratio.mild <- boxplot.stats(Female$RATIO, coef = 1.5)$out
Female.Ratio.ext <- boxplot.stats(Female$RATIO, coef = 3.0)$out

kable(mydata[mydata$RATIO %in% Female.Ratio.mild, ], caption = "Female RATIO mild outliers")
kable(mydata[mydata$RATIO %in% Female.Ratio.ext, ], caption = "Female RATIO extreme outliers")   #for these two, searched through the 'RATIO' column of 'mydata' to first match the mild, then the extreme outliers of Female abalones and list their rows with their associated variables in 'mydata'. There are 5 mild RATIO outliers and there is 1 extreme RATIO outlier for female abalones.


Infant.Ratio.mild <- boxplot.stats(Infant$RATIO, coef = 1.5)$out
Infant.Ratio.ext <- boxplot.stats(Infant$RATIO, coef = 3.0)$out

kable(mydata[mydata$RATIO %in% Infant.Ratio.mild, ], caption = "Infant RATIO mild outliers")
kable(mydata[mydata$RATIO %in% Infant.Ratio.ext, ], caption = "Infant RATIO extreme outliers")  #for these two, searched through the 'RATIO' column of 'mydata' to first match the mild, then the extreme outliers of Infant abalones and list their rows with their associated variables in 'mydata'. There are 7 mild RATIO outliers and there is 1 extreme RATIO outlier for infant abalones.


Male.Ratio.mild <- boxplot.stats(Male$RATIO, coef = 1.5)$out
Male.Ratio.ext <- boxplot.stats(Male$RATIO, coef = 3.0)$out  

kable(mydata[mydata$RATIO %in% Male.Ratio.mild, ], caption = "Male RATIO mild outliers")
kable(mydata[mydata$RATIO %in% Male.Ratio.ext, ], caption = "Male RATIO extreme outliers")    #for these two, searched through the 'RATIO' column of 'mydata' to first match the mild, then the extreme outliers of Male abalones and list their rows with their associated variables in 'mydata'. There are 5 mild RATIO outliers and 0 extreme RATIO outliers for male abalones.

```

**Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: The female, infant, and male abalone groups all have five or more mild RATIO outliers on the higher end. The female and infant abalone groups each have one extreme RATIO outlier. All infant RATIO outliers are mostly class A1 and two of class A2, all male RATIO outliers are from class A3, and female RATIO outliers are mostly of class A3, with one also from classes A2 and A5.***

-----

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS (Davies Section 14.3.2). There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}

CLASS <- factor(mydata$CLASS)
VOLUME <- mydata$VOLUME
WHOLE <- mydata$WHOLE

b1 <- ggplot(mydata, aes(CLASS, VOLUME)) + geom_boxplot()

b2 <- ggplot(mydata, aes(CLASS, WHOLE)) + geom_boxplot()


RINGS <- mydata$RINGS

g1 <- ggplot(mydata, aes(RINGS, VOLUME)) + geom_point()

g2 <- ggplot(mydata, aes(RINGS, WHOLE)) + geom_point()

grid.arrange(b1, b2, g1, g2, ncol=2, nrow=2)

```

**Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: I do not think these variables would perform well as predictors of age. First, the abalones' VOLUME and WHOLE differentiated by CLASS boxplots show wide ranges (whiskers) of most common values for classes A2 through A5. In class A1, 91 out of 108 are infant abalones, so A1 helps show common volume and whole values for some (less than a third) infants. However, 198 infant abalones are found in classes A2 and A3 for both volume and whole variables, and as shown, classes A2 through A5 have very similar volume and whole values. The higher the number for the class should represent older abalones, but if we took samples and knew only these variables (volume and whole weight), say for example, volumes of 150-300 cubic centimeters and whole weights of 40-100 grams, it would be extremely difficult to tell the age of the samples. Looking at the scatterplots of VOLUME and WHOLE versus RINGS, a large amount of abalones fall into the category of having between 5-15 rings as shown from the dense, filled in circles, so a sample abalone of volume 250 cubic cm. and whole weight of 50 grams can take on an age ranging anywhere from 6.5 to 16.5 years, potentially even older. These variables are not good predictors of age.***

-----

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
aggVolume <- aggregate(VOLUME~SEX+CLASS, data = mydata, mean)

aggShuck <- aggregate(SHUCK~SEX+CLASS, data = mydata, mean)

aggRatio <- aggregate(RATIO~SEX+CLASS, data = mydata, mean)

dim.matrix <- list(c("Female", "Infant", "Male"), c("A1", "A2", "A3", "A4", "A5"))

matrixVolume <- matrix(aggVolume[, 3], nrow = 3, ncol = 5, byrow = FALSE, dimnames = dim.matrix)
matrixShuck <- matrix(aggShuck[, 3], nrow = 3, ncol = 5, byrow = FALSE, dimnames = dim.matrix)
matrixRatio <- matrix(aggRatio[, 3], nrow = 3, ncol = 5, byrow = FALSE, dimnames = dim.matrix)

#Volume Matrix
kable(matrixVolume, digits = 2, caption = "$Volume")

#Shuck Matrix
kable(matrixShuck, digits = 2, caption = "$Shuck")

#Ratio Matrix
kable(matrixRatio, digits = 4, caption = "$Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
labels.sex <- c("Female", "Infant", "Male")

g1 <- ggplot(data = aggRatio, aes(x = CLASS, y = RATIO, group = SEX, 
        colour = SEX))+ geom_line()+ geom_point(size = 3)+ 
        ggtitle("Mean Ratio per Class")+ scale_color_hue(labels = labels.sex)

g2 <- ggplot(data = aggVolume, aes(x = CLASS, y = VOLUME, group = SEX, 
        colour = SEX))+ geom_line()+ geom_point(size = 3)+ 
        ggtitle("Mean Volume per Class")+ scale_color_hue(labels = labels.sex)

g3 <- ggplot(data = aggShuck, aes(x = CLASS, y = SHUCK, group = SEX, 
        colour = SEX))+ geom_line()+ geom_point(size = 3)+ 
        ggtitle("Mean Shuck per Class")+ scale_color_hue(labels = labels.sex)

grid.arrange(g1, g2, g3, newpage = TRUE)


```

**Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: The mean volume and mean shuck plots show that on average, infant abalones have the lowest shuck and volume values across all five classes while male and female are larger. However, questions are risen when considering the mean ratio of each sex per class. The ratios of the abalones are not very far off at all, with there being no distinction between each sex in class A1, and there being about a 0.01 difference in ratio between infant and adult abalones in classes A2 through A5. I think these plots show how much variability there is of adult (female and male) abalones' shuck and volume measurements. Male volumes range from 100 up to 450 cubic cm. and shuck weights range from 20 up to 60 grams, while infant volumes range from 75 up to 350 cubic cm. and shuck weights range from 10 up to 40 grams, when looking across all five classes. Female abalone measurements are higher values, but still show similar variability. ***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}

InfantAb <- subset(mydata, mydata$SEX == "I" & mydata$RINGS < 16)    #First, I wanted to separate Infant and Adult abalones, pulling from mydata only the "I" infants. Also only chose Infants under 16 rings.

AdultAb <- subset(mydata, (mydata$SEX == "F" | mydata$SEX == "M") & mydata$RINGS < 16)     #I did the same here, pulling "M" and "F" from mydata to get all Adult abalones.

box1 <- ggplot(InfantAb, aes(factor(InfantAb$RINGS), InfantAb$VOLUME)) + geom_boxplot() + ggtitle("Infant Volume | Rings") + labs(x = "Rings") + labs(y = "Volume") + ylim(0, 1100)

box2 <- ggplot(AdultAb, aes(factor(AdultAb$RINGS), AdultAb$VOLUME)) + geom_boxplot() + ggtitle("Adult Volume | Rings") + labs(x = "Rings") + labs(y = "Volume") + ylim(0, 1100)


box3 <- ggplot(InfantAb, aes(factor(InfantAb$RINGS), InfantAb$WHOLE)) + geom_boxplot() + ggtitle("Infant Whole weight | Rings") + labs(x = "Rings") + labs(y = "Whole weight") + ylim(0, 400)


box4 <- ggplot(AdultAb, aes(factor(AdultAb$RINGS), AdultAb$WHOLE)) + geom_boxplot() + ggtitle("Adult Whole weight | Rings") + labs(x = "Rings") + labs(y = "Whole weight") + ylim(0, 400)


grid.arrange(box1, box2, box3, box4, ncol = 2, nrow = 2)


```

**Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: These displays show that as rings increase (representing abalones growing) for both infant and adult abalones, volume and weight also increase, up to ring sizes of around 11, growth stabilizes and dips a bit from ringe size 11 to 15. When comparing the infant and adult displays, I notice first of all that each infant display shows less variability and smaller IQR's per ring group, while the adult displays show way more variability and a wider range for IQR's and longer whiskers per ring group. This is interesting and important to note that the adult displays have wide ranges of abalone whole weight and volume data that stretches enough to include the data of the infant measurements from the infant displays.***

-----

**Conclusions**

Please respond to each of the following questions (10 points total):

**Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: ***

**Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: Where was the sample collected from (what data frame) and how was it collected? What kind of random sample was it? Or was it nonrandom and why? Can you tell me about any known sampling or nonsampling errors (missing data, recording errors, processing, etc.)? Is the population normally distributed? ***

**Question 3)  (2 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: Difficulties of analyzing data derived from observational studies include the difficulty of determining how closely the observational sample represents the population, deciding whether to keep mild or extreme outliers or throw them away, ***
